---
title: "RepData_PeerAssessment1"
author: "Matt LeVasseur"
date: "Friday, March 06, 2015"
output: html_document
---
This assignment makes use of data from a single anonymous individual's personal
activity monitoring device, collecting data at 5 minute intervals throughout 
the day, from an anonymous individual during October-November, 2012.  

## Activity Steps Analysis

This analysis looks at:

- The average number of steps per day
- How steps change during the day 
- Whether step activity appears to change from weekday to weekend.

### Loading the Data

The datafile was forked from the course instructor's github repository: 
https://github.com/rdpeng/RepData_PeerAssessment1 

There are many fine ways of unzipping downloaded files, and this is a single
activity compared to the analysis. So I setup the script to be re-run several 
times, and chose to leave off the unzipping code so it is not run every time. 

Reading in and inspecting the .csv file from the working directory:

```{r readdf}

activity <- read.csv("activity.csv", header = TRUE)

```

The variables in this dataset are:

- **steps:** Number of steps taken in a 5-minute interval (missing values = NA)  
- **date:** The date the measurement was taken in YYYY-MM-DD format  
- **interval:** Identifier for the 5-minute interval of the measurement  

The dataset is a comma-separated-value (CSV) file with 17,568 observations.


### Average (Mean) Steps per Day

For this part, we are advised to ignore the missing values in the dataset.

**Calculate the total number of steps/day**

Using 'dplyr' to group and sum steps by date, for a table of total steps/day.
Data are divided into 61 unique days.

```{r}

library(dplyr)
activity_tbl <- tbl_df(activity)
group_date <- group_by(activity_tbl, date)
step_day <- summarize(group_date, sum(steps))

names(step_day) <- c("date" , "stepsum")   
                            # rename column 2 for manipulation and display
as.numeric(step_day$stepsum)# needed to produce histogram
step_day                    # to display step_day table explicitly in document

```

**Make a histogram of the total number of steps taken each day**

Making the grouping a seperate table to plot summary points into a histogram.

```{r hist1}

hist(step_day$stepsum, main= "Subject's Step/Day distribution", 
        xlab= "steps/day", col = "blue3")

```

**Calculate and report the mean and median of the total number of steps/day:**

-The subject's mean = 10,766.19 steps/day
-The median is similar = 10,765 steps/day

The similarity between these two values implies a somewhat symmetrical shape.

```{r stepavgs}

mean(step_day$stepsum, na.rm=TRUE)
median(step_day$stepsum, na.rm=TRUE)

```

### The average daily activity pattern

How do steps vary by *time interval* across *all days*? 

1. **Time series plot** (i.e. type = "l") of the 5-minute intervals (x-axis) 
and the average number of steps taken across all days (y-axis).

Still using dplyr:

```{r timeseries1}
# for this part, going to remove the observations where steps = NA
# this is necessary to be able to calculate means

    badobs <- is.na(activity$steps)
    activityok <- activity[!badobs, ] #subset of activity df, with only not = NA

    library(dplyr)
    interval_tbl <- tbl_df(activityok)      # need a tbl to work with dplyr

# steps/interval are single points along time, not a continuum
# create a time index for x-axis to plot avg steps one data point each    

    interval_tbl$interval <- as.factor(interval_tbl$interval)
    
# group data by interval factor
    group_interval <- group_by(interval_tbl, interval)
    step_interval <- summarize(group_interval, mean(steps))
    
# step_interval is a table of intervals summed over all days

# set the column names for display and to avoid '()'
names(step_interval) <- c("Interval" , "MeanSteps")

#plot will initially suppress x-axis, to insert Interval vector

plot(step_interval$MeanSteps, type="l", col="red", 
     main="Average steps per interval, all days", 
     xlab = "daily interval", ylab = "mean(steps)", 
     xaxt = "n")                    

#insert Interval vector for proper labels

axis(1, 1:288, labels = step_interval$Interval)
    
```

Here is the data table for the time series plot:

```{r}
# to display sample of table explicitly from which plot is made
step_interval[90:130, ]

```

2. **Which 5-minute interval, on average across all the days in the dataset,**
**contains the maximum number of steps?**

The maximum = interval 835 with a mean number of steps = 206.2, calculated 
below and and shown in the excerpt of the data table above.

```{r}
step_interval[which.max(step_interval$MeanSteps),]
```


### Imputing missing values

There are a number of days/intervals where there are missing values (coded NA). 
Missing days may introduce bias into some calculations or summaries of the data.

1. **Calculate and report the total number of missing values in the dataset** 
(i.e. the total number of rows with NAs)

- 2304 obs of steps recorded are NA
- 15,840 intervals have steps recorded, to include zero as an observed value

```{r}

summary(activity)

```


2 & 3. **Devise a strategy for filling in all of the missing values in the dataset**
For example, mean/median for that day, mean for that 5-minute interval, etc:

- **Imputing Strategy:** Use StepMean for the interval from 'step_interval' 
data table above. This strategy would be chosen if the interval is assumed
to be most influential, and most representative of missing data.

This strategy could lead to bias if the particular day turns out to be more
important, but is a reasonable strategy for this stage of the exploration, 
especially given the apparent influence shown in the time-series plot:

-**Code** showing how to create a new dataset equal to the original but missing data filled in, using this strategy:

```{r imputingloop}

imputed <- activity     # identical to original without disturbing original
names(imputed) <- c("Steps", "Date" , "Interval")
imputed$Interval <- as.factor(imputed$Interval)

interval_join <- left_join(imputed, step_interval, by ="Interval")

lastrow <- nrow(imputed)

#condition, if steps = NA for each row
# then replace with imputation strategy

for (i in 1:lastrow) {
    if(is.na(imputed[i,"Steps"]))   
        
        {imputed[i, "Steps"]  <-  interval_join[i, "MeanSteps"]
        }    
    }

# now the table 'imputed' has NA's replaced by interval means
```


4. ** Histogram of the total number of steps taken each day with imputed data**
and calculate the mean and median for total steps/day

Regrouping new data by date, as before but using the 'imputed' data table:

```{r}
library(dplyr)

imputed_date <- group_by(imputed, Date)
imputed_stepday <- summarize(imputed_date, sum(Steps))

names(imputed_stepday) <- c("Date","StepSum") # rename for manipulation, display

hist(imputed_stepday$StepSum, main= "Subject's Steps/Day with imputed data", 
        xlab= "steps/day", col = "red")

mean(imputed_stepday$StepSum, na.rm=TRUE)
median(imputed_stepday$StepSum, na.rm=TRUE)

imputed_stepday         # to display step_day table explicitly in document

```


For the histogram shown above:

**Calculate and report the mean and median total number of steps taken per day.** 

- Mean = 10766.19
- Median = 10766.19

**Do these values differ from the estimates from the first part of the assignment?**

Using my strategy, the mean is the same, the median has moved into line with the median.

**What is the impact of imputing missing data on the estimates of the total** 
**daily number of steps?**

The impact of imputing missing data with the strategy of using average interval
values adds to the number of total data points, particularly in the MIDDLE of the 
distribution; it accentuates the center of the data because we have added many
average values throughout the day.


###Differences in activity patterns between weekdays and weekends

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part **'imputed'**.

1. **Create a new factor variable in the dataset with two levels - "weekday"** 
**and "weekend" indicating whether a given date is a weekday or weekend day.**

```{r}
library(dplyr)

# currently 'imputed' has Date as a factor. Must convert to date type
imputed$Date <- as.Date(imputed$Date, "%Y-%m-%d")

# adding two columns, modifying the second to preserve weekday as well as factor
imputed <- mutate(imputed, DayofWeek=weekdays(imputed$Date))
imputed <- mutate(imputed, Weekend=weekdays(imputed$Date))

last <- nrow(imputed)

for (i in 1:last) {
    
    if(imputed[i,"DayofWeek"]== c("Saturday"))
        {imputed[i, "Weekend"] <- c("Weekend")}

    else if(imputed[i, "DayofWeek"]== c("Sunday"))
        {imputed[i, "Weekend"] <- c("Weekend")}
    
    else {imputed[i, "Weekend"] <- c("Weekday")} 
        }    

# To show samples of the resulting data table
imputed[1:10,]
imputed[1500:1510,]
imputed[1890:1900,]
```

2.**Make a panel plot containing a time series plot (i.e. type = "l") of the**
**5-minute interval (x-axis) and the average number of steps taken, averaged**
**across all weekday days or weekend days (y-axis).** 

See the README file in the GitHub repository to see an example of what this plot
should look like using simulated data. 

*Remember that this plot can be created using ANY plotting system*, 
it does not have to be LATTICE.

```{r}
library(lattice)
library(dplyr)

imputed$Weekend <- as.factor(imputed$Weekend)

# Must group by both factors, the interval and whether it is a weekend or not
imputed2 <- group_by(imputed, Interval, Weekend)
imputed_wknd <- summarize(imputed2, mean(Steps))

# rename for manipulation, display
names(imputed_wknd) <- c("Interval","Weekend", "MeanSteps") 

#separate dataset to make a single x axis label
xlabels <- filter(imputed_wknd, Weekend == "Weekday")

xyplot(MeanSteps ~ Interval | Weekend, data = imputed_wknd,
       layout= c(1,2), type="l", lty = 1, col="blue", xlab="Interval Index Number",
       scales=list(x = list(at=seq(from=0, to= 280, by = 20))))

```

It appears that the subject is able to walk more consistently through the 
daily intervals on weekends than on weekdays, which has a long spike early in
the day followed by extended periods of lower level activity.






